---
title: "combining_data"
author: "Sam Gartenstein"
date: "2025-04-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
path <- "data/processed data/individual"
```


### Demographic
```{r}
demographics_df <- read.csv(file.path(path, "demographics_df.csv"))
demographics_df
```

### Income
```{r}
income_df <- read.csv(file.path(path, "income_df.csv"))
income_df

```

### Education and Politic Variables
```{r}
educ_pol_df <- read.csv(file.path(path, "educ_pol_df.csv"))
educ_pol_df

```

## Examining Data

### Examining State Abbrevations

This is to ensure all the States are represented 

**Demographics**
```{r}
unique(demographics_df$State)
```

**Income**
```{r}
unique(income_df$State)
```


**Education and Politics**
```{r}
unique(educ_pol_df$State)
```

Since DC is not represented in the education and politics variable, we will have to omit DC from demographics and income.

### Examining Years

This is to ensure all years are represented 

**Demographics**
```{r}
unique(demographics_df$Year)
```

**Income**
```{r}
unique(income_df$Year)
```


**Education and Politics**
```{r}
unique(educ_pol_df$Year)
```

We need to get rid of 1990 through 1996 and 2014 and beyond in demographics.

### Additional Preprocessing

```{r}
demographics_df <- demographics_df %>%
  filter(State != "DC")

income_df <- income_df %>%
  filter(State != "DC")
```

**Double Checking DC is Removed**

```{r}
unique(demographics_df$State)
```

```{r}
unique(income_df$State)
```

### Removing Unecessary Years from Demographics 

```{r}
demographics_df <- demographics_df %>%
  filter(!(Year %in% 1990:1996 | Year %in% 2014:2019))
```

```{r}
unique(demographics_df$Year)

```

### Merging Dataframes 

```{r}
merged_df <- demographics_df %>%
  inner_join(income_df, by = c("State", "Year"))

merged_df
```

```{r}
merged_df <- merged_df %>%
  inner_join(educ_pol_df, by = c("State", "Year"))

merged_df
```

### Exporting 

```{r}
write.csv(merged_df, "data/processed data/merged/merged_df.csv", row.names = FALSE)

```


