---
title: "Demographic Variable Preprocessing"
author: "Sam Gartenstein"
date: "2025-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(tidyr)
library(dplyr)
library(purrr)
```

## Reading Data

```{r}
path <- "data/raw-data/demographics"
```

### Age Group Data

```{r}
agepop_data <- read_csv(file.path(path, "cdc_agepop_estimates_1990_2019.csv"))
head(agepop_data)
```
### Race Data

```{r}
racepop_data <- read_csv(file.path(path, "cdc_racepop_estimates_1990_2019.csv"))
head(racepop_data)
```

### Ethnicity Data

```{r}
ethpop_data <- read_csv(file.path(path, "cdc_ethpop_estimates_1990_2019.csv"))
head(ethpop_data)
```

## Preprocessing Functions

```{r}
# If the Notes column contains "Total", replace the value in the specified target column
# For example, "Age Group" has an empty cell for the total (not value), so this gives
# a label of "Total"
fill_total_note <- function(df, target_column) {
  df[[target_column]] <- ifelse(
    grepl("Total", df$Notes, ignore.case = TRUE),
    df$Notes,
    df[[target_column]]
  )
  return(df)
}
```


```{r}
# Function to drop unecessary variables
drop_vars <- function(df, vars_to_drop) {
  df[ , !(names(df) %in% vars_to_drop)]
}
```


```{r}
# Function to replace full name with abbreviations using R's built in
# state.name and state.abb vectors
# Manually changes District of Columbia to DC
abbreviate_states <- function(df, column_name = "State") {
  # Add DC manually
  full_names <- c(state.name, "District of Columbia")
  abbreviations <- c(state.abb, "DC")
  
  # Replace full state names with abbreviations
  df[[column_name]] <- ifelse(
    df[[column_name]] %in% full_names,
    abbreviations[match(df[[column_name]], full_names)],
    df[[column_name]]  # Leave unchanged if no match
  )
  
  return(df)
}
```


```{r}
# Function to rename "Yearly July 1st Estimates Code" to "Year"
rename_year_column <- function(df) {
  names(df)[names(df) == "Yearly July 1st Estimates"] <- "Year"
  return(df)
}
```

https://rlang.r-lib.org/reference/dot-data.html

https://dplyr.tidyverse.org/reference/mutate.html

```{r}

reshape_to_wide <- function(df, names_from_col) {
  df %>%
    dplyr::group_by(State, `State Code`, Year, .data[[names_from_col]]) %>%
    dplyr::summarise(Population = sum(Population, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(
      id_cols = c("State", "State Code", "Year"),
      names_from = all_of(names_from_col),
      values_from = "Population"
    ) %>%
    dplyr::arrange(`State Code`, Year)
}

```


## Preprocessing

### Replacing with Total

**Age Data**
```{r}
agepop_data <- fill_total_note(agepop_data, "Age Group")
head(agepop_data, 25)
```

**Race Data**
```{r}
racepop_data <- fill_total_note(racepop_data, "Race")
head(racepop_data, 25)
```


**Ethnicity Data**
```{r}
ethpop_data <- fill_total_note(ethpop_data, "Ethnicity")
head(ethpop_data, 25)
```



### Dropping Unecessary Columns

**Age Data**
```{r}
agepop_data <- drop_vars(agepop_data, c("Notes", "Yearly July 1st Estimates Code", "Age Group Code"))
head(agepop_data)
```


**Race Data**
```{r}
racepop_data <- drop_vars(racepop_data, c("Notes", "Yearly July 1st Estimates Code", "Race Code"))
head(racepop_data)
```

**Ethnicity Data**
```{r}
ethpop_data <- drop_vars(ethpop_data, c("Notes", "Yearly July 1st Estimates Code", "Ethnicity Code"))
head(ethpop_data)
```



### Renaming State Names to Abbreviations

**Age Data**
```{r}
agepop_data <- abbreviate_states(agepop_data, "State")
head(agepop_data)
```

**Race Data**
```{r}
racepop_data <- abbreviate_states(racepop_data, "State")
head(racepop_data)
```

**Ethnicity Data**
```{r}
ethpop_data <- abbreviate_states(ethpop_data, "State")
head(ethpop_data)
```

### Year 

**Age Data**
```{r}
agepop_data <- rename_year_column(agepop_data)
head(agepop_data, 25)
```

**Race Data**
```{r}
racepop_data <- rename_year_column(racepop_data)
head(racepop_data)
```

**Ethnicity Data**
```{r}
ethpop_data <- rename_year_column(ethpop_data)
head(ethpop_data)
```

```{r}
colnames(agepop_data)
```


### Reshaping Data

**Age**
```{r}
agepop_wide <- reshape_to_wide(agepop_data, "Age Group")
head(agepop_wide)
```

**Race**
```{r}
racepop_wide <- reshape_to_wide(racepop_data, "Race")
head(racepop_wide)
```


```{r}
racepop_wide
```

### Dropping Columns 

**Age**
```{r}

cols_to_drop <- c(
  "20-24 years", "25-29 years", "30-34 years", "35-39 years", "40-44 years",
  "45-49 years", "50-54 years", "55-59 years", "60-64 years", "65-69 years",
  "70-74 years", "75-79 years", "80-84 years", "85+ years", 
  "Show Totals: True", "NA"
)

agepop_wide <- agepop_wide %>%
  dplyr::select(-any_of(cols_to_drop)) %>%
  tidyr::drop_na()

agepop_wide

```

**Race**
```{r}
cols_to_drop <- c(
  "American Indian or Alaska Native", "Asian or Pacific Islander", "Black or African American",
  "Show Totals: True", "NA"
)

racepop_wide <- racepop_wide %>%
  dplyr::select(-any_of(cols_to_drop)) %>%
  tidyr::drop_na()

racepop_wide

```

### Calculating Percentages

**Age**

Calculating number 19 and under

```{r}
agepop_wide$`19_under` <- rowSums(
  agepop_wide[, c("< 1 year", "1-4 years", "5-9 years", "10-14 years", "15-19 years")],
  na.rm = TRUE
)

agepop_wide
```

```{r}
# Calcualting percentage of population under 19
agepop_wide$under19_pop_prct <- agepop_wide$`19_under` / agepop_wide$Total 
agepop_wide
```



```{r}
# Dropping remaining ages
cols_to_drop <- c(
  "< 1 year", "1-4 years", "5-9 years", 
  "10-14 years", "15-19 years", "19_under",
  "Total"
)

agepop_wide <- agepop_wide %>%
  dplyr::select(-any_of(cols_to_drop))

agepop_wide
```

**Race**

Calculating percentage White 

```{r}
# Calcualting percentage of white population
racepop_wide$white_pop_prct <- racepop_wide$White / racepop_wide$Total 
racepop_wide

```

```{r}
# Dropping uncessary columns 

cols_to_drop <- c(
  "Total", "White"
)

racepop_wide <- racepop_wide %>%
  dplyr::select(-any_of(cols_to_drop)) 

racepop_wide

```

```{r}
demographics_df <- dplyr::left_join(
  agepop_wide, 
  racepop_wide, 
  by = c("State", "Year")
)

demographics_df
```

```{r}
demographics_df <- demographics_df %>%
  dplyr::select(-`State Code.x`, -`State Code.y`)

demographics_df
```

```{r}
demographics_df <- demographics_df %>%
  arrange(State, Year)

demographics_df

```

### Exporting Data
```{r}
write.csv(demographics_df, "data/processed data/individual/demographics_df.csv", row.names = FALSE)

```



